---
title: "Week_1"
format: html
editor: visual
---

# Introduction

## Install R and R studio

R software: <https://cran.r-project.org/bin/windows/base/R-4.6.1-win.exe>

RStudio: <https://posit.co/downloads>

R bible:

Github:

## Project, folder organization and good practices

1.  Create a general folder for your research project.
2.  Split your folder into data, outputs, R script, papers (personal way - there is no recipe)
3.  Good practices: commenting the code - all information that you consider relevant. Outline (layout). Writing standard

## Interface R

1.  R Script
2.  Console
3.  Environment,
4.  Files, plots, packages, help

## Logical programming

Process thinking

1 + 1 x 2

```{r}
library(tidyverse)
a <- 1:10

a <- seq(1, 10, 1)

sum(sqrt(a))
a %>% 
  sqrt() %>%
  sum()


```

## Shortcut

**Ctrl + Enter** = run code

**Alt + -** = "\<-"

**Ctrl + Alt + i** = chunck

**Ctrl + Shift + m** = %>% (pipe)

**Ctrl + Shift + c** = add multiple comments


## Operators: calculator and booleans

1.  +, -, /, \*, %\*%, \^

2.  \> , \<, \>=, \<=, ==, !=, &, \|

```{r}
2 ^ 3

v = 2
m =3 

v * m

```

```{r}
3 >2

3 == 2

&

```


## Types of objects in R

Good names of object: short, clean, self explain, no space, no numbers, no special characters.

1.  Integer
2.  Numeric
3.  Character
4.  Logical
5.  Matrix
6.  Dataframe (Data structure board)
7.  Obs: vector (not a type)

```{r}
a = 1
a = as.integer(a)

b = 2.3
c = "Yuri"
c <- c("2", "3")

d = TRUE
e = matrix(c(1:4), ncol = 2)
f = data.frame(yield = c(80,90, 100),
               N_rate = c(20, 50, 80),
               Location = "Ashland",
               wheat = c(T, T, F))

```


## Function and packages

- What is a function?

- What is a function? To execute a sequence of operation

- Main functions

1.  sum()

2.  read_xlsx()

3.  summary()

4.  head(), tail()

5.  str()

6.  sqrt()

7.  table()

8.  seq()

9.  lenght()

1. Colnames()

```{r}
set.seed(10)
rnorm(10, mean = 1, sd = 2)
```

- What is a package? It's a set of functions

- How to install and use a package?

```{r}
# install.packages("")
# "package":: to use just specific function
# library()
```

- Main Packages

1.  base - base package (already installed)

2.  ggplot2 - data visualization

3.  dplyr - data analysis

4.  readxl - open excel file

## Function

- How to create a function?

Main structure

```{r}
new_function <- funtion(a, b, c){
  
  # command
  
  #return(result)
  
}
```

Arguments
Fixed set of objects


## Loops





# What is your goal?
