---
title: "Untitled"
author: "Hintz A."
date: "2026-08-17"
output: html_document
---

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction

In this workshop, we will introduce some basic concepts of data analysis and data organization in R.


# Packages

Packages provide additional functions and datasets that extend the capabilities of R.

For this workshop, we will use:

- agridat: agricultural datasets;
- tidyverse: a collection of packages for data manipulation, visualization, and analysis.

```{r, message=FALSE, warning=F}
library(agridat)
library(tidyverse)
```

# Exploring the data

## Selecting a Dataset

The agridat package contains many datasets related to agricultural experiments.

To see the datasets available in the package:

```{r}
data(package = "agridat")
```

For this workshop, we will use the sorghum dataset:
```{r}
dat <- agridat::adugna.sorghum

head(dat)
```
  

In practice, data are often stored in Excel or CSV files.

For example, an Excel file can be imported using:

```{r}
# dat <- read_excel("path/to/your_file.xlsx")
```

A CSV file can be imported using:

```{r}
# dat <- read_csv("path/to/your_file.csv")
```


Question: What are the observations in this dataset? What does each row represent?


## Structure of the Data

Before analyzing a dataset, we should understand its structure.

The str() function provides information about:

- the number of observations;
- the variables;
- the type of each variable;
- some example values.

```{r}
str(dat)
```

## Descriptive summary

The summary() function provides a quick descriptive summary of each variable.

```{r}
summary(dat)
```

## Checking missing values

In R, missing values are represented by NA.

We can count the number of missing values in each variable using:

```{r}
colSums(is.na(dat))
```

If a variable contains missing values, we need to consider how they should be handled before performing an analysis.

For example, when calculating a mean, we can use:

```{r}
mean(dat$yield, na.rm = TRUE)
```

The argument na.rm = TRUE tells R to remove missing values before calculating the mean.

## Frequency table

The table() function can be used to count combinations of categorical variables.

For example, we can examine how many observations are available for each combination of location and year:

```{r}
table(dat$loc, dat$year)
```


# Pipes

The pipe allows us to pass the result of one operation directly into another operation.

There are two commonly encountered pipe operators in R:

1.The native R pipe: |>
1. The magrittr pipe: %>%

For example:
```{r}
x <- c(1:10)

sqrt(x)

sum(sqrt(x))

sqrt(x) |> 
  sum()

```

The pipe makes it easier to read a sequence of operations from left to right.

# The Tidyverse   

The tidyverse is a collection of R packages designed to make data analysis easier and more consistent.

One of the most commonly used packages in the tidyverse is dplyr, which provides functions for manipulating data.

Some important dplyr functions are:

- select() — select columns;
- filter() — select rows;
- arrange() — reorder rows;
- mutate() — create or modify columns;
- rename() — rename columns;
- group_by() — define groups;
- summarise() — calculate summaries.



## Selecting Columns with select()

The select() function is used to select columns from a data frame.

First, we can look at the column names:

```{r}
colnames(dat)

dat |> 
  select(yield)

```


### Select Helpers

The select() function can also use helper functions.

  - starts_with()
  - ends_with()
  - contains(): 
  - matches(): 
  - num_range()
        
```{r}

#Select variables whose names start with a specific character or string.
dat |> select(starts_with("y"))

# Select variables whose names end with a specific character or string.
dat |> select(ends_with("loc"))

#Select variables whose names contain a specific character or string.
dat |> select(contains("e"))

#matches() uses a regular expression to identify column names.
#For example, to select variables whose names contain either "y" or "e":

dat |> select(matches("y|e"))


#num_range() is useful when variables follow a numerical naming pattern.
#for example, suppose we have variables named G18, G19, and G20:

dat_gen <- dat |> 
  mutate(value = 1) |> 
  pivot_wider(
    names_from = gen,
    values_from = value,
    values_fill = 0
)

dat_gen |> select(num_range("G", 18:20))

```


### Selecting Columns Without Tidyverse

The same operation can be performed using base R.

```{r}
dat2 <- dat[, c("env", "loc")]

#We can also select columns by their position:
dat2 <- dat[, c(3, 6)]
```

Both approaches work, but select() is often easier to read when working with more complicated data manipulation pipelines.


## Filtering rows with filter()

The filter() function selects observations that satisfy a logical condition.


```{r}
# Filtering a Range

# Option 1: "between" function
dat |> 
  filter(between(yield, 500, 700))

# Option 2: logical operations
dat |> 
  filter(yield >= 500 & yield <= 700)


# Filtering one location

dat |> filter(loc == "Kobo")

# Multiple values: 
##The %in% operator is useful when we want to select several categories.

dat |> filter(gen %in% c("G01", "G02", "G03"))

# Combining conditions 
#AND

dat |> filter(loc == "Kobo" & year == 2001)

# OR
dat |> filter(loc == "Kobo" | loc == "Melkassa")

# Negation
dat |> filter(loc != "Kobo")
```


## Ordering Rows with arrange()

The arrange() function is used to reorder observations.

```{r}
# Reorder the data frame rows according to the highest-to-low yield order
dat |> arrange(desc(yield))

# Sorting ascending, for comparison
dat |> arrange(yield)

```


## Creating variables with mutate()

The mutate() function creates new variables or modifies existing variables.

```{r}
# Convert yield from kg/ha to t/ha
dat |> mutate(yield_t = yield / 1000)
```


mutate() becomes particularly useful when combined with group_by().

For example, we can calculate the mean yield for each location and then calculate the deviation of each observation from its location mean:

```{r}
dat |> 
  group_by(loc) |> 
  mutate(
    loc_mean_yield = mean(yield, na.rm = TRUE),
    yield_dev = yield - loc_mean_yield
  ) |> 
  ungroup()
```

The ungroup() function removes the grouping after the calculation.


We can use case_when() to create categorical variables based on numerical values.

```{r}
# mutate() + case_when() to create a categorical column from a numeric one
dat |> 
  mutate(
    yield_class = case_when(
      yield < 500  ~ "low",
      yield < 1000 ~ "medium",
      TRUE         ~ "high"
    )
  )
```


## Renaming Variables

The rename() function changes the name of a variable.

```{r}
dat |> rename(genotype = gen, location = loc)
```


## Using across()

The across() function allows us to apply the same operation to multiple variables.


```{r}
dat |> 
  mutate(across(where(is.factor), as.character)) |> 
  str()

dat |> 
  mutate(across(c(gen, env, loc), as.character)) |> 
  str()

```

across() becomes especially useful when we need to apply the same transformation to many columns.


## Grouping and summarising

The combination of group_by() and summarise() is one of the most useful tools for exploratory data analysis.

group_by() defines the groups we want to analyze.

summarise() calculates one or more summary statistics for each group.

```{r}

dat |> 
  group_by(loc) |> 
  summarise(mean_loc = mean(yield),
            min_yield = min(yield),
            max_yield = max(yield))


```


```{r}
dat |> 
  group_by(loc, year) |> 
  summarise( mean_yield = mean(yield, na.rm = TRUE), 
             n = n(),
             .groups = "drop" )
```



# Practice exercises

The exercises are based on the data `yield.xlsx`, so first read the file and save it in the object `df`.

1. Verify how many observations and variables are in the dataset and what are the categories of the variables.

2. Select just the column `location`, `plot`, `block`, `trt`, `system`, `wheat_yield_bu_acre_13.5`,
`soy_yield_bu_acre_13`

3. Rename the variables that contain the wheat and soybean yield.

4. Verify if there are any NAs in the dataset.

5. Filter only the SW system and calculate the mean wheat yield for each location.

6. Calculate the mean and standard deviation for each system for wheat and soybean yield.


 




