---
title: "Week_1"
format: pdf
editor: visual
---

# Introduction

Welcome! This first session covers the basics you need to learning R: installing R and RStudio, organizing a project, understanding the RStudio interface, and the core building blocks of the R language (objects, functions, operators, and loops).

## Install R and R studio

- **R** is the programming language and engine that actually runs your code.
- **RStudio** is the interface (IDE) that makes working with R easier — script editor, console, plots, help.

1.  Install R: <https://cran.r-project.org/bin/windows/base/R-4.6.1-win.exe>

2.  Install RStudio: <https://posit.co/downloads>

3.  R bible: <https://r4ds.hadley.nz/>, *R for Data Science*

## Project and folder organization — good practices

1.  Create one general folder per research project
2.  Split the folder into subfolders. A common structure (there's no single "correct" recipe, adapt to your workflow):
    - `data/` — raw and processed data
    - `scripts/` (or `R/`) — your `.R` / `.qmd` scripts
    - `outputs/` — figures, tables, model results
    - `papers/` or `docs/` — literature, manuscripts, notes
3.  Good coding practices:
    - **Comment your code**. Future-you (and your collaborators) will thank you.
    - **Use an outline/layout** — section headers (`#`, `##`, `###` in comments or Markdown) keep long scripts navigable.
    - **Use a consistent naming/writing standard** — e.g. always `snake_case` for object names, consistent indentation, consistent spacing around `<-`.

## Interface R

RStudio is organized into four main panes:

1.  **Source / R Script** (top-left) — where you write and save code.
2.  **Console** (bottom-left) — where code actually runs and have the error messages.
3.  **Environment / History** (top-right) — shows the objects currently loaded in memory, and a record of past commands.
4.  **Files / Plots / Packages / Help** (bottom-right) — file browser, plot viewer, package manager, and documentation.

## Logical programming

Programming means telling the computer a precise sequence of steps — think of it as a recipe. Before writing code, it helps to think through the logic first: what are the inputs, what steps happen, what is the output?

## Keyboard shortcuts

| Shortcut                 | Action                                      |
|--------------------------|---------------------------------------------|
| **Ctrl + Enter**         | Run current line / selection                |
| **Alt + -**              | Insert `<-`                                 |
| **Ctrl + Alt + I**       | Insert a new code chunk (R Markdown/Quarto) |
| **Ctrl + Shift + M**     | Insert pipe `%>%`                           |
| **Ctrl + Shift + C**     | Comment/uncomment selected lines            |
| **Ctrl + L**             | Clear the console                           |
| **Ctrl + Shift + Enter** | Run entire chunk                            |

## Operators

**Arithmetic:** `+`, `-`, `/`, `*`, `%*%` (matrix multiplication), `^` (power)

**Relational:** `>`, `<`, `>=`, `<=`, `==` (equal to), `!=` (not equal to)

**Logical:** `&` (and), `|` (or), `!` (not)

```{r}
3 ^ 3
2 * 3
2 + 3
2 - 3
2 / 3
```

```{r}
3 > 2
3 == 2

TRUE & FALSE   # AND -> both must be true
TRUE | FALSE   # OR  -> at least one must be true
!TRUE          # NOT -> flips the value
```

# Types of objects in R

Good object names: short, clear, self-explanatory, no spaces, avoid starting with numbers, avoid special characters (use `_` instead).

1.  **Integer** — whole numbers
2.  **Numeric** (double) — decimal numbers
3.  **Character** — text/strings
4.  **Logical** — `TRUE` / `FALSE`
5.  **Vector** — a sequence of values\
6.  **Matrix** — a 2-dimensional structure
7.  **Data frame** — a table with observation and variables
8.  **List** — a flexible container that can hold objects of different types and sizes together

```{r}
a <- 1
a <- as.integer(a)
class(a)

b <- 2.3
class(b)

c_var <- c("2", "3")   # character vector
class(c_var)

d <- TRUE
class(d)

e <- matrix(c(1:4), ncol = 2)
e

f <- data.frame(
  yield    = c(80, 90, 100),
  N_rate   = c(20, 50, 80),
  Location = "Ashland",
  wheat    = c(TRUE, TRUE, FALSE)
)
f
str(f)
```

# Function and packages

**What is a function?** A set of instructions, packaged under a name, that executes a sequence of operations on some input and returns a result. Instead of retyping the same steps every time, you call the function.

**Common built-in functions:**

1.  `sum()`
2.  `read_xlsx()` / `read.csv()`
3.  `summary()`
4.  `head()`, `tail()`
5.  `str()`
6.  `sqrt()`
7.  `table()`
8.  `seq()`
9.  `length()`
10. `colnames()`
11. `print()`

```{r}
c(1,2,3,4,5,6,7,8,9,10)
c(1:10)
seq(1, 10, 1)
```

**What is a package?** A collection of functions (and sometimes data) written by someone else, that you can install and load to expand what R can do without needing anything else.

**Installing and using a package:**

```{r}
# install.packages("dplyr")   # run once, to download the package
# dplyr::filter()              # use one function without loading the whole package
# library(dplyr)               # load the whole package for the session
```

**Main packages:**

1.  **base** — pre-installed, core R functions
2.  **ggplot2** — data visualization
3.  **dplyr** — data manipulation/analysis
4.  **readxl** — reading Excel files
5.  **tidyr** — reshaping data (long/wide formats)

## Writing your own function

General structure:

```{r, eval: false}

new_function <- function(a, b, c) {

  # commands / calculations

  return(result)
}
```

Example:

```{r}
my_mean <- function(a, b) {
  result <- (a + b) / 2
  return(result)
}

my_mean(2, 4)
```

## Loops

Loops repeat a task a set number of times, or until a condition is no longer true.

- `for()` — repeat a fixed, known number of times
- `while()` — repeat as long as a condition stays true

Structure:

```{r}
n = 10
for(i in 1:n){
  print(i)
}
```

```{r}
i= 1
while(i < n) {
  
  print(i)
  
  i = i + 1
}

```

# Conditions

# Practice exercises

1.  Create a vector `n_rates` with the values 0, 40, 80, 120, 160 (kg N/ha).
```{r}

n_rates = c(0,40,80,120,160)

```

2.  Create a data frame with three columns: `plot` (1 to 5), `n_rate` (`n_rates` above), and `yield` (make up five plausible wheat yield values).

```{r}

data.frame(plot = c(1:5), 
           n_rate = n_rates,
           yield = c(20,30,40,50,60))

```


3.  Write a function `kg_to_lb(kg)` that converts kilograms to pounds (1 kg = 2.2046 lb) and test it on your `yield` column.


4.  Write a `for` loop that prints "Plot X received Y kg N/ha" for each row of your data frame.



git checkout --ours index.qmd
git checkout --ours week_1.qmd
git checkout --ours week_1.html